---
title: "Demo Rmd file"
author: "Jane Doe"
date: "2024-06-20"
output: html_document
---

```{r setup, include=FALSE}
# this is a standard "setup" chunk usually found at the top of Rmd files,
# often used for setting options, loading files, and importing libraries
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Section 1

## Subsection A

Here's some ordinary text. You can use Markdown syntax to add more features, e.g. here's a [link](https://markdownguide.org/cheat-sheet), here's some **bold text**, and here's some `inline code`. You can also add images, footnotes, blockquotes, and more. See linked cheat sheet above for more.

 1. Lists are also east to add!
 2. Here's a second item.
 3. You can even add sublists:
    - Here's a sublist with bullets.
    - Another bullet?

## Subsection B

You can easily incorporate R code into an Rmd file, with outputs and plots that auto-update. Here's an example code chunk named "chunk1".

```{r example-chunk, fig.height=5, fig.align="center"}
data <- c(3, 6, 6, 2, 4, 1, 5)
mean(data)
hist(data)
```

You can even refer to R objects inside text, e.g. the sample mean and standard deviation are `r mean(data)` and `r sd(data)`.

# Section 2

Here's a second section.

<!-- comments in an Rmd file must use HTML-style syntax -->
